Timid tides or tall tails? \\ Investigating the outer reaches of the Sculptor and Ursa Minor dwarf galaxies

### 2025-05-19 thesis/acronyms.tex

@@

\newacronym
[ description ={Core collapse supernovae. Massive star explosions. {\sc Ccsne} produce many elements including \gls{alpha}, \gls{fepeak}, and r and \gls{sproc}.}
]
{cc}{ccsne}{core collapse supernovae}

\newacronym
[description={Asymptotic giant branch stars. The {\sc agb} phase is the last phase of \glspl{lowmass}, before stars become white dwarfs. Produces C, N, and heavy \gls{sproc} elements. See \ref{sec:stel_evo}.}.]
{agb}{agb}{asymptotic giant branch}

\newacronym
[description={Type Ia Supernovae. Exploding white dwarfs. Produces \gls{fepeak} and has a long delay time.}]
{ia}{sne ia}{supernovae type Ia}

\newacronym
[description={Galactic chemical evolution.}]
{gce}{gce}{galactic chemical evolution}

\newacronym
[description={Star formation history.}]
{sfh}{sfh}{star formation history}

\newacronym
[description={Single stellar population. A group of stars formed all at the same time.}]
{ssp}{ssp}{single stellar population}

@@

\newacronym
[
description={Damped Lyman-alpha system. {\sc dla}s are clouds of gas from the 
early universe which are observed through their absorption of quasar spectra. The name 
comes from the strong Lyman-alpha lines ($1216\AA$) due to H absorption.}
]
{dla}{dla}{damped Lyman-alpha systems}

\newcommand{\cc}{\gls{cc}}
\newcommand{\Cc}{\Gls{cc}}
\newcommand{\agb}{\gls{agb}}
\newcommand{\ia}{\gls{ia}}
\newcommand{\sfh}{\gls{sfh}}
\newcommand{\dla}{\gls{dla}}
\newcommand{\ssp}{\gls{ssp}}
\newcommand{\imf}{\gls{imf}}
\newcommand{\gce}{\gls{gce}}
\newcommand{\Gce}{\Gls{gce}}
\newcommand{\apogee}{\gls{apogee}}

@@

\newglossaryentry{multizone}{name={multi-zone},
    description={A chemical evolution model where a galaxy is divided into several zones, each with different stars, gas, and properties. }
}

\newglossaryentry{onezone}{name={one-zone},
    description={A chemical evolution model where the gas is all the same composition, i.e. neglecting spatial variations.}
}

\newglossaryentry{alpha}{name={$\alpha$-element},
    description={Elements which are made up of $\alpha$-particles (He nuclei) throuch the triple-$\alpha$ process (see Eq.~\ref{eq:triple_alpha}). Essentially light, even-numbered elements like O, Mg, and Na.}
}

\newglossaryentry{fepeak}{name={Fe-peak elements},
    description={Fe and nearby elements, produced in \gls{cc} and \gls{ia}.}
}

\newglossaryentry{sproc}{name={s-process elements},
    description={Elements produced through slow neutron captures, typically in \agb\ stars.}
}

\newglossaryentry{rgb}{name={red giant branch},
    description={Red giant branch stars are stars that have completed hydrogen core burning and have expanded in size. See Appendix \ref{sec:stel_evo}.}
}

\newglossaryentry{tdu}{name={third dredge up},
    description={Third dredge up occurs inside \gls{agb} stars. During each thermal pulse, material is {\it dredged up} from the core, changing the chemical abundances of the stellar atmosphere. (While nominally called {\it third dredge up}, there are typically several third dredge ups.). See Appendix \ref{sec:stel_evo}.}
}

\newglossaryentry{fdu}{name={first dredge up},
    description={First dredge up occurs when a \gls{lowmass} enters the \gls{rgb} phase. Material from the core is brought to the surface, increasing N and decreasing C abundances. See Appendix \ref{sec:stel_evo}.}
}

\newglossaryentry{hbb}{name={hot bottom burning},
    description={Hot bottom burning occurs inside \agb\ stars. The base of the convective envelope becomes hot enough for CNO burning to initiate.}
}

\newglossaryentry{subgiant}{name={subgiant},
    description={A star in the process of leaving the main sequence and becoming a \gls{rgb}.}
}

\newglossaryentry{imfave}{name={\gls{imf}-averaged},
    description={Averaged over the initial-mass function ({\sc imf}). The {\imf}-weighted yield is the mass of the newly produced element divided by the mass of star formation for a single stellar population (see \ssp).}
}

\newglossaryentry{dtd}{name={delay time distribution},
    description={The distribution in time of when an element is produced 
    after a star formation event.}
}

\newglossaryentry{massloading}{name={mass loading},
    description={The strength of outflows relative to star formation. See also $\eta$. }
}

\newglossaryentry{lowalpha}{name={low-$\alpha$},
    description={The low-$\alpha$ sequence, as described by Eq.~\ref{eq:high_alpha}. }
}

\newglossaryentry{highalpha}{name={high-$\alpha$},
    description={The high-$\alpha$ sequence, as described by Eq.~\ref{eq:high_alpha}. }
}

\newglossaryentry{lowmass}{name={low-mass star},
    description={Stars with masses $\lesssim 8\,M_\odot$ which end life as a white dwarf. }
}

\newglossaryentry{highmass}{name={high-mass star},
    description={Stars with masses $\lesssim 8\,M_\odot$, which end as a neutron star, black hole, or supernovae.}
}

\newglossaryentry{insideout}{name={{\it insideout}},
    description={Our fiducial star formation history. The rate of star formation is highest towards the center of the galaxy and at earlier times. See Eq.~\ref{eq:inside_out}.}
}

\newglossaryentry{cno}{name={CNO cycle},
    description = { A proton-fusion cycle which occurs in \gls{rgb} stars 
        consisting of a chain of proton captures releasing a He nucleus ($\alpha$-particle) and energy. See Eq.~\ref{eq:cno_cycle}.
}
}

>>>>

% Ordinary abbreviations
\newcommand{\kms}{{\rm km\,s^{-1}}}
\newcommand{\kmsdeg}{{\rm km\,s^{-1}\,deg^{-1}}}

@@

@@

@@

### 2025-05-19 thesis/appendix.tex

>>>>

\section{Density profile tests}\label{density-profile-tests}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_density_methods_extra.pdf}}
\caption[Density profiles]{Density profiles for various assumptions for
Sculptor. PSAT is our fiducial 2-component J+24 sample, circ is a
2-component bayesian model assuming circular radii, simple is the series
of simple cuts described, bright is the sample of the brightest half of
stars (scaled by 2), DELVE is a sample of RGB stars (background
subtracted and rescaled to
match).}\label{fig:sculptor_observed_profiles}
\end{figure}

Note that a full rigorous statistical analysis would require a
simulation study of injecting dwarfs into Gaia and assessing the
reliability of various methods of membership and density profiles. This
is beyond the scope of this thesis.

\begin{verbatim}
SELECT TOP 1000
       *
FROM delve_dr2.objects
WHERE 11 < ra
and ra < 19
and -37.7 < dec
and dec < -29.7
\end{verbatim}

\subsection{An MCMC model}\label{an-mcmc-model}

We additionally consider a model which solves for the density relative
to the background in each annulus. This model is specified as \[
\log\Sigma_i \sim {\rm U}(-12, 6)
\] where the stars are statically placed into each bin. This model
solves for the density in each bin independently.

\section{Velocity modelling and
comparisons}\label{velocity-modelling-and-comparisons}

Here, we describe in additional detail, our methods and comparisons for
RV modelling between studies.

Savage-Dickey calculated Bayes factor using Silverman-bandwidth KDE
smoothed samples from posterior/prior.

\begin{table*}[t]
\centering
\caption{Summary of velocity measurements and derived properties. sestito+2023a number of members depends on spatial model used.}
\begin{tabular}{lllllllll}
\toprule
 & Study & Instrument & Nspec & Nstar & Ngood & Nmemb & $\delta v_{\rm med}$ & $R_{\rm xmatch}$/arcmin\\
\midrule
Scl & combined &  & 8945 & 2280 & 2034 & 1918 & 0.9 & \\
 & tolstoy+23 & FLAMES & 3311 & 1701 & 1522 & 1480 & 0.65 & –\\
 & sestito+23a & GMOS & 2 & 2 & 2 & 2 & 13 & –\\
 & walker+09 & MMFS & 1818 & 1522 & 1417 & 1329 & 1.8 & 3\\
 & APOGEE & APOGEE & 5082 & 253 & 102 & 98 & 0.6 & –\\
UMi & combined &  & 4714 & 1225 & 1148 & 831 & 2.3 & \\
 & sestito+23b & GRACES & 5 & 5 & 5 & 5 & 1.8 & –\\
 & pace+20 & DEIMOS & 1716 & 1538 & 829 & 682 & 2.5 & 1\\
 & spencer+18 & Hectoshell & 1407 & 970 & 596 & 406 & 0.9 & ?\\
 & APOGEE & APOGEE & 9500 & 279 & 37 & 32 & 0.9 & –\\
\bottomrule
\end{tabular}
\end{table*}

\begin{table*}[t]
\centering
\caption[Sculptor RV fits]{\label{tbl:scl_rv_mcmc}MCMC fits for different RV datasets for Sculptor among 3 different models. }
\label{tbl:scl_rv_mcmc}
\begin{tabular}{lllllllllll}
\toprule
study & mean & sigma & $\partial \log\sigma / \partial \log R$ & $\partial v_z / \partial x$ (km/s/deg) & $\theta_{\rm grad} / ^{\circ}$ & $\hat R$ & $n_{\rm eff}$ & $\log B_2/B_1$ & WAIC & LOO\\
\midrule
all &  &  &  &  &  &  &  &  &  & \\
 & $111.18\pm0.23$ & $9.71\pm0.17$ & - & - & - &  &  & 0 &  & \\
 & $111.19 \pm 0.23$ & $9.68\pm0.17$ & - & $4.3\pm1.3$ & $-149_{-13}^{+17}$ &  &  & -1.7 &  & \\
 & $111.16\pm0.23$ & $9.73\pm0.17$ & $0.056\pm0.021$ & - & - &  &  & -0.9 &  & \\
tolstoy+23 &  &  &  &  &  &  &  &  &  & \\
 & $111.3 \pm 0.3$ & $9.80 \pm 0.18$ & - & - & - &  &  & 0 &  & \\
 & $111.3\pm0.3$ & $9.78\pm0.18$ & – & $4.3\pm1.4$ & $-154_{-13}^{+19}$ &  &  & -1.4 &  & \\
 & $111.2 \pm 0.3$ & $9.74\pm0.19$ & $0.085 \pm 0.023$ & – & – &  &  & -4.5 &  & \\
walker+09 &  &  &  &  &  &  &  &  &  & \\
 & $111.0\pm0.3$ & $9.57\pm0.21$ & – & – & – &  &  & 0 &  & \\
 & $111.1\pm0.3$ & $9.53\pm0.21$ & - & $5.2_{-1.6}^{+1.8}$ & $-134_{-16}^{+23}$ &  &  & -2.4 &  & \\
 & $111.0\pm0.3$ & $9.61\pm0.21$ & $0.03\pm0.03$ & – & – &  &  & +1.6 &  & \\
apogee &  &  &  &  &  &  &  &  &  & \\
 & $109.9\pm0.8$ & $8.3\pm0.6$ & – & – & – &  &  & – &  & \\
 & $109.9\pm0.8$ & $8.3\pm0.6$ & – & $6\pm3$ & $-151_{-36}^{+44}$ &  &  & +0.4 &  & \\
 & $109.9\pm0.8$ & $8.3\pm0.7$ & $0.05\pm0.08$ & – & – &  &  & +1.1 &  & \\
\bottomrule
\end{tabular}
\end{table*}

\begin{table*}[t]
\centering
\caption[Ursa Minor RV fits]{\label{tbl:umi_rv_mcmc}MCMC fits for UMi velocity dispersion. }
\label{tbl:umi_rv_mcmc}
\begin{tabular}{lllllll}
\toprule
study & mean & sigma & $\hat R$ & $n_{\rm eff}$ & $\log bf_{\rm sigma}$ & $\log bf_{\rm grad}$\\
\midrule
all & $-245.9\pm0.3$ & $8.76\pm0.24$ &  &  & +1.5 & +2.2\\
pace & $-244.5\pm0.4$ & $9.1\pm0.3$ &  &  & +0.2 & +1.1\\
spencer & $-246.9\pm0.4$ & $8.8\pm0.3$ &  &  & +1.8 & -0.3\\
apogee & $-248.2\pm1.6$ & $9.0_{-1.1}^{+1.3}$ &  &  & +0.8 & +0.8\\
\bottomrule
\end{tabular}
\end{table*}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_rv_scatter.pdf}}
\caption[Scl velocity sample]{RV members of Sculptor plotted in the
tangent plane coloured by corrected velocity difference from mean
\(v_z - \bar v_z\) . The black ellipse marks the half-light radius in
fig.~\ref{fig:scl_selection}. The black and green arrows mark the proper
motion (PM, GSR frame) and derived velocity gradient (rot) vectors (to
scale).}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_vel_gradient_scatter.pdf}}
\caption[Scl velocity gradient]{The corrected LOS velocity along the
best fit rotational axis. RV members are black points, the systematic
\(v_z\) is the horizontal grey line, blue lines represent the
(projected) gradient from MCMC samples, and the orange line is a rolling
median (with a window size of 50).}
\end{figure}

\section{Numerical Convergence}\label{numerical-convergence}

### 2025-05-19 thesis/sculptor.tex

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250415112706417.png}}
\caption{image-20250415112706417}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250415113122385.png}}
\caption{image-20250415113122385}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250415113133353.png}}
\caption{image-20250415113133353}
\end{figure}

>>>>

\section{Abstract}\label{abstract}

\section{Introduction}\label{introduction}

Sculptor (Scl) is one of the first discovered dwarf galaxies of the
Milky Way (Shapley 1938; only preceded by the SMC and LMC!). As a
classical dwarf spheroidal, Scl is relatively bright and compact.

Since the initial discovery of Scl, most authors have noted that Scl has
a slight ellipticity (\(0.36\)), often attributed to tidal effects.
However, this does not align with the absolute proper motion or orbital
path of the galaxy.

While Scl has a relatively large pericentre (greater than 50kpc),
\citet{sestito+2023a} detect that the galaxy likely has an excess of
stars in the outskirts (past about 60 arcminutes). This excess is
perhaps one of the clearest indications that Scl may be affected by
tides. Here, our goal is to determine if under a \(\Lambda\) CMD
paradigm with DM-only simulations, tidal effects of the Milky Way (or
LMC) may indeed be consistent with these observations, or if these
observations may reveal instead a extended stellar ``halo'' or second
component of the galaxy---illustrating a complex formation for the
galaxy.

Theoretical work on Sculptor

\begin{itemize}
\tightlist
\item
  \citet{battaglia+2008}
\item
  \citet{iorio+2019}
\item
  \citet{amorisco+zavala+deboer2014}
\item
  \citet{battaglia+2008}
\item
  \citet{breddels+2013}
\item
  \citet{breddels+helmi2013}
\item
  \citet{richardson+fairbairn2014}
\item
  \citet{SFW2017}
\item
  \citet{innanen+papp1979}
\item
  \citet{wilkinson+2002}
\item
  \citet{yang+2025}: chemical evolution in Scl.
\item
  \citet{skuladottir+2024}; \citet{skuladottir+2021};
  \citet{lee+jeon+bromm2024}, pop III high res spectro?
\item
  \citet{wang+2024a} hydrosim of dwarf galaxies like sculptor
\item
  \citet{tang+2023}: apogee modeling of Scl.
\item
  \citet{pandey+west2022} chem evo OMEGA and isotopes.
\item
  \citet{an+koposov2022}: distance gradients?
\item
  \citet{kawata+2006}, 2pop origins?
\end{itemize}

Observational work on Scl

\begin{itemize}
\tightlist
\item
  \citet{sestito+2023a}
\item
  \citet{westfall+2006} wide degree survey for extended structure.
\item
  \citet{tolstoy+2023}, \citet{arroyo-polonio+2023},
  \citet{arroyo-polonio+2024}
\item
  \citet{eskridge1988}, \citet{eskridge1988a}, \citet{eskridge1988b}
\item
  \citet{coleman+dacosta+bland-hawthorn2005}
\item
  \citet{DQ1994}
\item
  \citet{WMO2009}
\item
  \citet{IH1995},
\item
  \citet{munoz+2018}: Using Megacam to derive density profiles and
  structural properties of many dwarf spheroidal galaxies.
\item
  \citet{kirby+2009}
\item
  \citet{martinez-vazquez+2015}, \citet{pietrzynski+2008}
\item
  \citet{grebel1996}
\item
  \citet{barbosa+2025}: Using DECam to derive narrowband photometric
  metallicity gradient and search for metal poor stars in Scl.
\end{itemize}

Future ideas:

\begin{itemize}
\tightlist
\item
  \citet{evslin2016}: measuring ellipticities of halos w TMT.
\end{itemize}

@@

\begin{figure}
\centering
\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250415112706417.png}
\caption[Effect of LMC on Sculptor's Orbit]{Orbits of Sculptor with and
without the LMC. The potential is that of \citet{vasiliev+2021} in each
case.}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250415113122385.png}
\caption[Selected orbits of Sculptor with an LMC]{Selected orbits of
Sculptor in the MW frame. \textbf{TODO}: include LMC distance or frame,
maybe xyz instead?}
\end{figure}
